---
title: "Comparing axiom systems"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparing axiom systems}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6,
                      fig.height = 4)
```

```{r setup}
library(pgt)
```

## Which axiom system for bad outputs?

How to model a technology that produces both a good and a bad output is
an open question in the productivity literature. The 2021 symposium in
the *Journal of Productivity Analysis* set the weak-disposability school
against the by-production school, with the materials-balance view running
through both. The choice of axiom changes the efficiency scores, so a
result that holds under one system need not hold under another. `pgt`
implements the competing systems on one data object and reports how far
their conclusions agree.

The models split into two families. The materials-balance models enforce
$u'x_l - v y_l \ge b_l$:

- `wgd`: weak-G-disposability (Rodseth 2025).
- `mb_cost`: the materials-balance cost model (Coelli et al. 2007),
  reporting the technical and environmental-allocative components of
  $EE = TE \times EAE$.

The reference models come from the competing systems:

- `byprod`: the by-production intersection technology (Murty, Russell and
  Levkoff 2012), reporting output efficiency, emission efficiency and
  their graph average.
- `wd`: weak disposability (Kuosmanen 2005), which imposes no materials
  balance.

## Fitting each model

```{r}
data(steeldemo)
tech <- pgt_tech(
  x = steeldemo[, c("coal_coke", "other_fuel", "raw_material", "flux")],
  y = steeldemo$production,
  b = steeldemo$emissions,
  v = 0.01467,
  group = steeldemo$route,
  id = steeldemo$plant
)

fits <- lapply(c("wgd", "byprod", "mb_cost", "wd"), function(m) {
  pgt(tech, model = m)
})
names(fits) <- c("wgd", "byprod", "mb_cost", "wd")
sapply(fits, function(f) round(median(f$results$efficiency, na.rm = TRUE), 3))
```

The by-production and materials-balance measures need not agree, because
they reduce a different quantity: `byprod` contracts the observed
emission over the emission-causing inputs, while `mb_cost` minimises the
total material inflow.

## The comparison harness

`compare_models()` fits the efficiency-scored models and reports rank
agreement and the overlap of their worst-performer sets.

```{r}
cmp <- compare_models(tech, models = c("wgd", "byprod", "mb_cost", "wd"))
cmp
```

The Spearman matrix shows how closely the models rank the plants. A high
correlation means the choice of axiom moves the scores but preserves the
ordering; a low correlation warns that the ranking itself depends on the
axiom. The `bottom_q_overlap` column reports the share of the reference
model's worst-quartile plants that each model also places in its worst
quartile, the plants a regulator would target first.

```{r, fig.alt = "Efficiency score profiles across four models"}
plot(cmp)
```

Reading the plot: plants are ordered along the horizontal axis by their
weak-G-disposability score, and each line is one model. Where the lines
track together the axiom choice is immaterial; where they cross, a plant's
relative standing depends on which system is used.
