---
title: "Introduction to pgt"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to pgt}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6,
                      fig.height = 4)
```

```{r setup}
library(pgt)
```

## The materials-balance principle

A pollution-generating technology turns material inputs into a good
output and an unavoidable bad output. The pollutant content of the
inputs has to end up somewhere: embodied in the product, removed by
abatement, or emitted. `pgt` builds efficiency models on that accounting
identity,
$$u'x_l - v y_l \ge b_l,$$
where $x_l$ are the material inputs of unit $l$, $u$ their pollutant
flow coefficients (for example tonnes of CO2 potential per unit of
input), $y_l$ the good output, $v$ the pollutant retained in the
product, and $b_l$ the emissions. The gap $u'x_l - v y_l - b_l$ is the
pollutant that leaves the process other than through the measured
emission.

Existing R packages handle undesirable outputs by transforming the data
(the Seiford-Zhu translation in `deaR`) or by imposing weak
disposability and directional distances (`nonparaeff`, `Benchmarking`).
Neither enforces the materials balance as a constraint. `pgt` does, and
puts the competing axiom systems behind one interface.

## Building a technology

The constructor bundles the data with the flow coefficients. The shipped
`steeldemo` data is a synthetic panel of steel plants on two production
routes, with emissions in CO2 units.

```{r}
data(steeldemo)
tech <- pgt_tech(
  x = steeldemo[, c("coal_coke", "other_fuel", "raw_material", "flux")],
  y = steeldemo$production,
  b = steeldemo$emissions,
  v = 0.01467,
  group = steeldemo$route,
  id = steeldemo$plant
)
tech
```

The inputs already sit in pollutant-potential units, so the default
$u = 1$ applies to each; `v = 0.01467` is the carbon retained per unit of
crude steel.

## Auditing the identity before estimation

The weak-G-disposability model gives every DMU a self-reference only when
its own materials balance holds. `mb_check()` reports the closure gap per
DMU and flags the accounts where more pollutant leaves than enters.

```{r}
mb <- mb_check(tech)
head(as.data.frame(mb))
attr(mb, "n_violations")
```

An infeasible weak-G-disposability program always belongs to a DMU with a
strictly negative gap, so this audit locates any estimation problem in
advance.

## Estimating environmental efficiency

The weak-G-disposability model minimises the bad output subject to output
and input feasibility, the peer emission envelope, and the DMU's own
materials-balance cap. The score is $b^*_l / b_l \in (0, 1]$.

```{r}
fit <- pgt(tech, model = "wgd")
summary(fit)
```

The envelope model frees the inputs and reduces the program to the
convex lower envelope of the $(y, b)$ scatter. Its self-reference is
always feasible.

```{r}
fit_env <- pgt(tech, model = "envelope")
summary(fit_env)
```

## Shadow prices and abatement cost

The output-constraint dual measures the marginal emission content of the
good output along the frontier. `mac_curve()` turns it into a marginal
abatement cost curve: reducing the bad output by one unit costs
$1 / \eta_l$ units of the good output, or $p / \eta_l$ at output price
$p$.

```{r}
head(shadow_prices(fit))
mac <- mac_curve(fit, price = 550)
plot(mac)
```

## Decomposing across production routes

With a technology group, `pgt_decompose()` splits environmental
efficiency into a within-group and a technology-gap component. The
envelope decomposition is an exact identity, Total $=$ WR $\times$ TGR,
where WR is within-route reallocation and TGR the technology-gap ratio of
Battese, Rao and O'Donnell (2004).

```{r}
dec <- pgt_decompose(tech, type = "envelope")
summary(dec)
```

## A worked example on real data

The `frontier` package ships a balanced panel of Philippine rice farms
with a physical fertiliser input (`NPK`, kilograms of active ingredient)
and a physical output (`PROD`, tonnes of paddy). Treating the nitrogen
carried by the fertiliser as the pollutant potential gives a nitrogen
materials balance. The nitrogen fraction below is illustrative
(urea-equivalent, about 0.46 of the active ingredient).

```{r, eval = requireNamespace("frontier", quietly = TRUE)}
data("riceProdPhil", package = "frontier")
d8 <- riceProdPhil[riceProdPhil$YEARDUM == 8, ]
uN <- 0.46
rice <- pgt_tech(
  x = as.matrix(d8[, c("AREA", "LABOR", "NPK", "OTHER")]),
  y = d8$PROD,
  b = uN * d8$NPK,
  u = c(AREA = 0, LABOR = 0, NPK = uN, OTHER = 0),
  v = 0,
  id = as.character(d8$FMERCODE)
)
rice_fit <- pgt(rice, model = "wgd")
summary(rice_fit)
```

Here environmental efficiency is the scope to cut fertiliser nitrogen
while holding rice output at its observed level. The other vignettes
cover model validation (`vignette("replication", "pgt")`), the
cross-axiom comparison (`vignette("comparing-axioms", "pgt")`) and
productivity measurement with inference
(`vignette("productivity", "pgt")`).
