---
title: "Productivity change and inference"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Productivity change and inference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6,
                      fig.height = 4)
```

```{r setup}
library(pgt)
```

## Productivity change with a bad output

The Malmquist-Luenberger index (Chung, Färe and Grosskopf 1997) measures
productivity change when part of the output is undesirable, by scoring
each unit on a directional distance function (Chambers, Chung and Färe
1996) that expands the good output and contracts the bad output
together. The adjacent-period version can be infeasible across periods
(Aparicio, Pastor and Zofío 2013); the global
index of Oh (2010) pools every period into one reference technology,
which removes the infeasibility and makes the index circular. `pgt_ml()`
computes it under the pollution-generating technology and splits it into
efficiency change and best-practice change,
$GML = EC \times BPC$, with $GML > 1$ signalling growth.

This estimator is experimental: the global Malmquist-Luenberger index
under a materials-balance technology is not yet settled in the
literature, and the default technologies do not re-impose the per-DMU
materials-balance cap on the projected point.

The `technology` argument selects the reference technology.
`technology = "wd"` imposes weak disposability of the bad output in the
Kuosmanen (2005) form; this is the technology under which Chung, Färe
and Grosskopf (1997) and Oh (2010) define the index, so it is the
faithful comparator. The default `"wgd"` and the `"envelope"` option
treat the bad output as reducible to the peer emission envelope and are
exploratory companions.

### A constructed panel

The panel below has a frontier that improves between the two periods:
the same output is produced with 15 percent less emission.

```{r}
set.seed(1)
n <- 15
d1 <- data.frame(id = 1:n, y = runif(n, 5, 10), x = runif(n, 8, 12),
                 b = runif(n, 2, 6))
d2 <- data.frame(id = 1:n, y = d1$y * 1.03, x = d1$x, b = d1$b * 0.85)
d <- rbind(cbind(d1, period = 1), cbind(d2, period = 2))
panel <- pgt_tech(x = d[, "x", drop = FALSE], y = d$y, b = d$b,
                  period = d$period, id = d$id)
ml <- pgt_ml(panel)
summary(ml)
```

The median index exceeds 1, and best-practice change carries it: the
frontier moved out while each unit kept its position relative to the
frontier, so efficiency change stays near 1.

### A real rice panel

The Philippine rice panel in the `frontier` package spans eight years.
Treating fertiliser nitrogen as the pollutant potential gives a
nitrogen-based productivity index.

```{r, eval = requireNamespace("frontier", quietly = TRUE)}
data("riceProdPhil", package = "frontier")
r <- riceProdPhil
uN <- 0.46
rice_panel <- pgt_tech(
  x = as.matrix(r[, c("AREA", "LABOR", "NPK", "OTHER")]),
  y = r$PROD,
  b = uN * r$NPK,
  u = c(AREA = 0, LABOR = 0, NPK = uN, OTHER = 0),
  v = 0,
  period = r$YEARDUM,
  id = as.character(r$FMERCODE)
)
rice_ml <- pgt_ml(rice_panel)
summary(rice_ml)
```

## Subsampling inference

Efficiency scores are point estimates from a finite sample of DMUs.
`boot_pgt()` gauges their sampling variability by re-solving each unit
against random subsamples of the reference set (m out of n, without
replacement), the resampling scheme with the firmest footing for
boundary estimators (Politis and Romano 1994; Politis, Romano and Wolf
1999; Kneip, Simar and Wilson 2008). `boot_pgt()` covers cross-sectional
efficiency scores only; no inference is currently available for the
`pgt_ml()` index. The example takes the first 60 plants of the synthetic
`steeldemo` panel.

```{r}
data(steeldemo)
steel60 <- steeldemo[1:60, ]
tech <- pgt_tech(
  x = steel60[, c("coal_coke", "other_fuel", "raw_material", "flux")],
  y = steel60$production, b = steel60$emissions, v = 0.01467,
  group = steel60$route, id = steel60$plant
)
bt <- boot_pgt(tech, model = "wgd", B = 40, seed = 1)
head(bt$per_dmu)
```

Each unit is evaluated against the subsample frontier alone; the unit
itself is not appended to the subsample. Replicates in which its
programme is infeasible are dropped and counted in `per_dmu$n_ok`, and
the point estimate can lie above its own upper endpoint, because the
interval targets the true score, which the estimate overestimates.

The intervals follow the subsampling recipe: the subsample scores are
recentred at the point estimate and rescaled by the relative
convergence rate $(m/L)^\kappa$, so the interval extends downward from
the estimate, matching the direction of the frontier bias (a smaller
reference sample can only move the estimated frontier inward). The
intervals remain heuristic: resampling theory for data envelopment
analysis covers the radial and directional technologies, not the
materials-balance programs here, so $\kappa$ borrows the DEA rate as an
approximation. Before relying on the intervals, check that their width
is stable in the subsample size (Simar and Wilson 2011).

```{r}
boot_pgt_sensitivity(tech, model = "wgd", B = 25,
                     m_grid = c(15, 25, 40), seed = 1)
```

A width that changes little across the grid supports the chosen
subsample size; a strong trend means the size, not the data, is driving
the interval.

When the technology carries groups, `boot_pgt()` also reports intervals
for the group mean scores. These are a descriptive stability band, not
supported inference: the replicate statistic averages all group members
against a common subsample frontier, so replicate variation reflects
movement of the frontier only, and the per-DMU convergence rate is
reused for the mean (Kneip, Simar and Wilson 2015 give the theory for
aggregates of efficiency scores).

<!-- coverage-table -->

## References

Aparicio, J., Pastor, J. T., & Zofío, J. L. (2013). On the inconsistency
of the Malmquist-Luenberger index. *European Journal of Operational
Research*, 229(3), 738-742. doi:10.1016/j.ejor.2013.03.031

Chambers, R. G., Chung, Y., & Färe, R. (1996). Benefit and distance
functions. *Journal of Economic Theory*, 70(2), 407-419.
doi:10.1006/jeth.1996.0096

Chung, Y. H., Färe, R., & Grosskopf, S. (1997). Productivity and
undesirable outputs: A directional distance function approach. *Journal
of Environmental Management*, 51(3), 229-240. doi:10.1006/jema.1997.0146

Kneip, A., Simar, L., & Wilson, P. W. (2008). Asymptotics and consistent
bootstraps for DEA estimators in nonparametric frontier models.
*Econometric Theory*, 24(6), 1663-1697. doi:10.1017/S0266466608080651

Kneip, A., Simar, L., & Wilson, P. W. (2015). When bias kills the
variance: Central limit theorems for DEA and FDH efficiency scores.
*Econometric Theory*, 31(2), 394-422. doi:10.1017/S0266466614000413

Kuosmanen, T. (2005). Weak disposability in nonparametric production
analysis with undesirable outputs. *American Journal of Agricultural
Economics*, 87(4), 1077-1082. doi:10.1111/j.1467-8276.2005.00788.x

Oh, D.-h. (2010). A global Malmquist-Luenberger productivity index.
*Journal of Productivity Analysis*, 34(3), 183-197.
doi:10.1007/s11123-010-0178-y

Politis, D. N., & Romano, J. P. (1994). Large sample confidence regions
based on subsamples under minimal assumptions. *The Annals of
Statistics*, 22(4), 2031-2050. doi:10.1214/aos/1176325770

Politis, D. N., Romano, J. P., & Wolf, M. (1999). *Subsampling*.
Springer, New York. doi:10.1007/978-1-4612-1554-7

Simar, L., & Wilson, P. W. (2011). Inference by the m out of n bootstrap
in nonparametric frontier models. *Journal of Productivity Analysis*,
36(1), 33-53. doi:10.1007/s11123-010-0200-4
