---
title: "Productivity change and inference"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Productivity change and inference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6,
                      fig.height = 4)
```

```{r setup}
library(pgt)
```

## Productivity change with a bad output

The Malmquist-Luenberger index measures productivity change when part of
the output is undesirable, by scoring each unit on a directional distance
that expands the good output and contracts the bad output together. The
adjacent-period version can be infeasible across periods; the global
index of Oh (2010) pools every period into one reference technology,
which removes the infeasibility and makes the index circular. `pgt_ml()`
computes it under the pollution-generating technology and splits it into
efficiency change and best-practice change,
$GML = EC \times BPC$, with $GML > 1$ signalling growth.

This estimator is experimental: the global Malmquist-Luenberger index
under a materials-balance technology is not yet settled in the
literature. When the materials-balance cap does not bind, the index
coincides with the standard weak-disposability version.

### A constructed panel

The panel below has a frontier that improves between the two periods:
the same output is produced with 15 percent less emission.

```{r}
set.seed(1)
n <- 15
d1 <- data.frame(id = 1:n, y = runif(n, 5, 10), x = runif(n, 8, 12),
                 b = runif(n, 2, 6))
d2 <- data.frame(id = 1:n, y = d1$y * 1.03, x = d1$x, b = d1$b * 0.85)
d <- rbind(cbind(d1, period = 1), cbind(d2, period = 2))
panel <- pgt_tech(x = d[, "x", drop = FALSE], y = d$y, b = d$b,
                  period = d$period, id = d$id)
ml <- pgt_ml(panel)
summary(ml)
```

The median index exceeds 1, and best-practice change carries it: the
frontier moved out while each unit kept its position relative to the
frontier, so efficiency change stays near 1.

### A real rice panel

The Philippine rice panel in the `frontier` package spans eight years.
Treating fertiliser nitrogen as the pollutant potential gives a
nitrogen-based productivity index.

```{r, eval = requireNamespace("frontier", quietly = TRUE)}
data("riceProdPhil", package = "frontier")
r <- riceProdPhil
uN <- 0.46
rice_panel <- pgt_tech(
  x = as.matrix(r[, c("AREA", "LABOR", "NPK", "OTHER")]),
  y = r$PROD,
  b = uN * r$NPK,
  u = c(AREA = 0, LABOR = 0, NPK = uN, OTHER = 0),
  v = 0,
  period = r$YEARDUM,
  id = as.character(r$FMERCODE)
)
rice_ml <- pgt_ml(rice_panel)
summary(rice_ml)
```

## Subsampling inference

Efficiency scores are point estimates from a finite sample of DMUs.
`boot_pgt()` gauges their sampling variability by re-solving each unit
against random subsamples of the reference set (m out of n, without
replacement), the resampling scheme with the firmest footing for
boundary estimators.

```{r}
data(steeldemo)
steel60 <- steeldemo[1:60, ]
tech <- pgt_tech(
  x = steel60[, c("coal_coke", "other_fuel", "raw_material", "flux")],
  y = steel60$production, b = steel60$emissions, v = 0.01467,
  group = steel60$route, id = steel60$plant
)
bt <- boot_pgt(tech, model = "wgd", B = 40, seed = 1)
head(bt$per_dmu)
```

The intervals follow the subsampling recipe: the subsample scores are
recentred at the point estimate and rescaled by the relative
convergence rate $(m/L)^\kappa$, so the interval extends downward from
the estimate, matching the direction of the frontier bias (a smaller
reference sample can only move the estimated frontier inward). The
intervals remain heuristic: resampling theory for data envelopment
analysis covers the radial and directional technologies, not the
materials-balance programs here, so $\kappa$ borrows the DEA rate as an
approximation. Before relying on the intervals, check that their width
is stable in the subsample size.

```{r}
boot_pgt_sensitivity(tech, model = "wgd", B = 25,
                     m_grid = c(15, 25, 40), seed = 1)
```

A width that changes little across the grid supports the chosen
subsample size; a strong trend means the size, not the data, is driving
the interval.
