---
title: "Replicating published results"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Replicating published results}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(pgt)
```

The estimators are validated against the numerical examples printed in
their source papers. These replications run as unit tests on every check;
this vignette reproduces them in the open.

## Rodseth (2025), Table 2: minimal controlled emissions

The `pigfarms` data is Table 1 of Rodseth (2025): five pig-finishing
farms with feed and piglets as material inputs, meat as the good output,
and controlled, uncontrolled and abatement emission columns for
nitrogen. The uncontrolled emission equals controlled plus abatement,
so the uncontrolled column is the pollutant carrier ($u = 1$,
$v = 0$).

```{r}
data(pigfarms)
pigfarms
```

```{r}
tech <- pgt_tech(
  x = pigfarms[, c("uncontrolled", "labor", "capital")],
  y = pigfarms$meat,
  b = pigfarms$controlled,
  u = c(1, 0, 0),
  a = pigfarms$abatement,
  id = pigfarms$farm
)
fit <- pgt(tech, model = "wgd")
data.frame(farm = fit$results$id, b_star = fit$results$b_star)
```

The minimal controlled emissions are $(16, 16, 16, 20, 16)$, which match
the paper's Table 2.

## Rodseth (2025), Table 3: the directional representation

The factorially determined multi-output model (Eq. 13, abatement fixed at
each farm's own level) jointly expands the good output and contracts the
bad output. Its good-output coefficient $v = 7/6$ follows from farms A
and B, which share inputs: their meat differs by 3 and their uncontrolled
emissions by 3.5. Each farm's feed coefficient closes its account to the
uncontrolled column.

```{r}
v <- 7 / 6
u_feed <- (pigfarms$uncontrolled + v * pigfarms$meat) / pigfarms$feed
tech3 <- pgt_tech(
  x = pigfarms[, c("feed", "piglet", "labor", "capital")],
  y = pigfarms$meat,
  b = pigfarms$controlled,
  u = cbind(u_feed, 0, 0, 0),
  v = v,
  a = pigfarms$abatement,
  id = pigfarms$farm
)
fdmo <- pgt(tech3, model = "fdmo")
r3 <- fdmo$results
data.frame(
  farm = r3$id,
  good_eff = round(r3$good_eff, 3),
  bad_eff = round(r3$bad_eff, 3),
  gross = round(r3$gross, 3),
  maximal_y = round(r3$maximal_y, 3)
)
```

Farms A, B and D are on the frontier (gross inefficiency 0). Farm C
expands its good output by 3 and cuts its bad output by 3.5, a gross
inefficiency of 6.5; these figures reproduce Table 3 exactly. Farm E's
good-output
expansion rounds to the paper's 0.8; its bad-output figure is
$v$ times the good-output figure, the ratio the materials balance forces
whenever uncontrolled emission equals controlled plus abatement.

## Murty, Russell and Levkoff (2012): the by-production example

Their Example 1 is a constant-returns technology with one input, one good
output $y$ and one bad output $z$, solved analytically in the paper.

```{r}
mrl <- pgt_tech(
  x = matrix(c(1, 1, 1, 2, 2), ncol = 1, dimnames = list(NULL, "x")),
  y = c(2, 3 / 2, 2 / 3, 3, 2),
  b = c(4, 1, 2, 5, 3),
  u = 1, polluting = "x",
  id = paste0("DMU", 1:5)
)
bp <- pgt(mrl, model = "byprod", returns = "crs")
rb <- bp$results
data.frame(id = rb$id, output_eff = round(rb$output_eff, 4),
           emission_eff = round(rb$efficiency, 4), fgl = round(rb$fgl, 4))
```

For DMU3 the output efficiency is $1/3$, the emission efficiency $1/2$
and the graph measure $5/12$; for DMU2 they are $3/4$, $1$ and $7/8$.
These match the analytic scores in the paper.

## Coelli, Lauwers and Van Huylenbroeck (2007): material-flow coefficients

Coelli et al. (2007) measure environmental efficiency against a
materials-balance cost model and report the phosphorus contents used
(their footnote 16): feed 0.0124, piglet 0.0117 and meat 0.0117 kg P per
kg. The cost model decomposes environmental efficiency as
$EE = TE \times EAE$.

```{r}
farms <- data.frame(
  feed = c(200, 210, 190, 205),
  piglet = c(20, 22, 18, 21),
  meat = c(100, 100, 95, 102)
)
tech_c <- pgt_tech(
  x = as.matrix(farms[, c("feed", "piglet")]),
  y = farms$meat,
  b = 0.0124 * farms$feed + 0.0117 * farms$piglet - 0.0117 * farms$meat,
  u = c(feed = 0.0124, piglet = 0.0117),
  v = 0.0117
)
mbc <- pgt(tech_c, model = "mb_cost", returns = "crs")
rc <- mbc$results
data.frame(id = rc$id, EE = round(rc$efficiency, 4),
           TE = round(rc$te, 4), EAE = round(rc$eae, 4))
all.equal(rc$efficiency, rc$te * rc$eae)
```

The identity $EE = TE \times EAE$ holds to machine precision.
